Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-01-08 paper/paper.tex
update paper intro

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets,
which include the size of the maximum independent sets, the number and enumeration of independent sets and maximal independent sets of a given size.
% which include the size of the maximum independent sets, the number of independent sets and maximal independent sets of a given size, and enumeration of independent sets and maximal independent sets of a given size.
%Although the computational complexity inevitably scales exponentially with treewidth of the graph,
%our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
%One can easily find a good contraction order for the tensor network representation of a graph $G$ with thousands of vertices. 
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the tree width of the problem graph.
Our framework can be easily extended to compute properties of other problems such as the cut size, coloring, and maximal cliques, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the overlap gap property on three regular graphs.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting and hard computational problems concerning various properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$. 
One of the independent set properties that many computational complexity theorists are most interested in is to find a maximum independent set (MIS) and the size of the MIS, $\alpha(G) \equiv \max_{I}|I|$,
not only because it has a wide range of applications in scheduling, logistics, wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}, 
but also because it is a well-known NP-complete problem~\cite{Hastad1996} that can be mapped from many other important combinatorial optimization problems in polynomial time such as the 3-satisfiability problem, the maximum clique problem, and the minimum vertex cover problem~\cite{Moore2011}.
However, a real world application often requires knowing more properties about independent sets,
which include, for example, the number and enumeration of independent sets or maximal independent sets of a given size.
Some of these properties are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022} \red{replace with experiment paper when ready};
they can, for example, be used to understand phase transitions, to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the overlap gap property~\cite{Gamarnik2013, Gamarnik2019}.
Meanwhile, finding the number of independent sets of a given size is equivalent to computing the coefficients of the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}.
%\blue{The primary interest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.}
%Finding $\alpha(G)$ exactly is a hard computational problem, and it is NP-hard even to approximate $\alpha(G)$ within $|V|^{1-\epsilon}$ ~\cite{Hastad1996},
%meaning there is unlikely a polynomial time algorithm to find an independent set with size $\alpha(G)/|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
%Naive exhaustive search for an MIS requires a computational time of $O\left(2^{|V|} \right)$.
%There is immense interest in finding better algorithms to solve the MIS problem, not only because it has 
%a wide range of applications in scheduling, logistics, wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}, 
%but also because it is a well-known NP-complete problem that can be mapped from many other important combinatorial optimization problems in polynomial time such as the 3-satisfiability problem, the maximum clique problem, and the minimum vertex cover problem~\cite{Moore2011}.

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig1.pdf}
    \caption{Benchmark results for computing different properties of independent sets of a random three regular graph with different tensor element types.
    The time in these plots only includes tensor network contraction, without taking the contraction order finding and just in time compiling time into account.
    Legends are properties, algebra and devices that we used in the computation, one can find the corresponding computed property in \Tbl{tbl:generictypes}.
    (a) time and space complexity versus the number of vertices for the benchmarked graphs.
    (b) The computing time for calculating the MIS size and for counting of the number of independent sets (ISs), the number of MISs, and the number of independent sets having size $\alpha(G)$ and $\alpha(G)-1$.
    (c) The computing time for calculating the independence polynomials with different approaches.
    (d) The computing time for configuration enumeration, including the enumeration of all independent set configurations, a single MIS configuration, all MIS configurations, all independent set configurations having size $\alpha(G)$ and $\alpha(G)-1$,  with or without bounding the enumeration space.
    \red{suggestions for the figure: 1. the legend seems to overlap the lines, especially the large sub-figure. Not sure what's a better way to include the legend there. 2. I see you guys don't typically use markers. We typically use markers to distinguish the different lines, as well as using the color. Especially, for some journals, there is a physical non-color copy, so some requires the figures to be understandable without colors.}
    }
    \label{fig:benchmark}
\end{figure}

>>>>

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets,
which include the size of the maximum independent sets, the number and enumeration of independent sets of a given size.
% which include the size of the maximum independent sets, the number of independent sets and maximal independent sets of a given size, and enumeration of independent sets and maximal independent sets of a given size.
%Although the computational complexity inevitably scales exponentially with treewidth of the graph,
%our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
%One can easily find a good contraction order for the tensor network representation of a graph $G$ with thousands of vertices. 
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the tree width of the problem graph.
Our framework can be easily extended to compute properties of other problems such as the cut size, coloring, and maximal cliques, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the overlap gap property on three regular graphs.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting and hard computational problems concerning various properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$. 
One of the independent set properties that many computational complexity theorists are most interested in is to find a maximum independent set (MIS) and the size of the MIS, $\alpha(G) \equiv \max_{I}|I|$,
not only because it has a wide range of applications in scheduling, logistics, wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}, 
but also because it is a well-known NP-complete problem~\cite{Hastad1996} that can be mapped from many other important combinatorial optimization problems in polynomial time such as the 3-satisfiability problem, the maximum clique problem, and the minimum vertex cover problem~\cite{Moore2011}.
Some other properties, for example, the number and enumeration of independent sets or maximal independent sets of a given size are euqally important in many practical applications, however, are not studied enough.
Some of them are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022} \red{replace with experiment paper when ready};
they can, for example, be used to understand phase transitions, to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the overlap gap property~\cite{Gamarnik2013, Gamarnik2019}.
Meanwhile, finding the number of independent sets of a given size is equivalent to computing the coefficients of the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}.
%\blue{The primary interest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.}
%Finding $\alpha(G)$ exactly is a hard computational problem, and it is NP-hard even to approximate $\alpha(G)$ within $|V|^{1-\epsilon}$ ~\cite{Hastad1996},
%meaning there is unlikely a polynomial time algorithm to find an independent set with size $\alpha(G)/|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
%Naive exhaustive search for an MIS requires a computational time of $O\left(2^{|V|} \right)$.
%There is immense interest in finding better algorithms to solve the MIS problem, not only because it has 
%a wide range of applications in scheduling, logistics, wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}, 
%but also because it is a well-known NP-complete problem that can be mapped from many other important combinatorial optimization problems in polynomial time such as the 3-satisfiability problem, the maximum clique problem, and the minimum vertex cover problem~\cite{Moore2011}.

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig1.pdf}
    \caption{Benchmark results for computing different properties of independent sets of a random three regular graph with different tensor element types.
    The time in these plots only includes tensor network contraction, without taking the contraction order finding and just in time compiling time into account.
    Legends are properties, algebra and devices that we used in the computation, one can find the corresponding computed property in \Tbl{tbl:generictypes}.
    (a) time and space complexity versus the number of vertices for the benchmarked graphs.
    (b) The computing time for calculating the MIS size and for counting of the number of independent sets (ISs), the number of MISs, and the number of independent sets having size $\alpha(G)$ and $\alpha(G)-1$.
    (c) The computing time for calculating the independence polynomials with different approaches.
    (d) The computing time for configuration enumeration, including the enumeration of all independent set configurations, a single MIS configuration, all MIS configurations, all independent set configurations having size $\alpha(G)$ and $\alpha(G)-1$,  with or without bounding the enumeration space.
    }
    \label{fig:benchmark}
\end{figure}
